### 2024-04-30 paper.tex
new bib name

\title{Bip32-Ed25519++: A More Secure Hierarchical Deterministic Wallet Standard for Ed25519 Keys}

\author{Chris Peikert\cite{Author1}, Yared Afework\cite{Author2}}

\newcommand{\abstractText}{\noindent

}

%%%%%%%%%%%%%%%%%
% Configuration %
%%%%%%%%%%%%%%%%%

\documentclass[12pt, a4paper, twocolumn]{article}
\usepackage{xurl}
\usepackage[super,comma,sort&compress]{natbib}
\usepackage{abstract}
\renewcommand{\abstractnamefont}{\normalfont\bfseries}
\renewcommand{\abstracttextfont}{\normalfont\small\itshape}
\usepackage{amsmath} % Add this line to define the \text command
\usepackage{titlesec}
\usepackage[skip=10pt]{parskip}

\titleformat*{\section}{\large\bfseries}
\titleformat*{\subsection}{\normalsize\bfseries}

%%%%%%%%%%%%%%
% References %
%%%%%%%%%%%%%%

% If changing the name of the bib file, change \bibliography{test} at the bottom
\begin{filecontents}{sources.bib}

@misc{Author1,
  author       = "Peikert, Chris",
  howpublished = "\url{mailto:cpeikert@umich.edu}",
}

@misc{Author2,
  author       = "Efrem Afework, Yared",
  howpublished = "\url{mailto:yared.efremafework@algorand.foundation}",
}

@article{BIP32-Ed25519,
  author  = "Khovratovich, Dmitry, and Law, Jason",
  title   = "BIP32-Ed25519 Hierarchical Deterministic Keys over a Non-linear Keyspace",
  year    = "2017",
  journal = "IEEE European Symposium on Security and Privacy Workshops (EuroS&PW)",
  note    = "\url{https://ieeexplore.ieee.org/abstract/document/7966967}",
}

\end{filecontents}

% Any configuration that should be done before the end of the preamble:
\usepackage{hyperref}
\hypersetup{colorlinks=true, urlcolor=blue, linkcolor=blue, citecolor=blue}

\begin{document}

%%%%%%%%%%%%
% Abstract %
%%%%%%%%%%%%

\twocolumn[
  \begin{@twocolumnfalse}
    \maketitle
    \begin{abstract}
      \abstractText
      Implementing BIP32-style HD wallets for Ed25519 keys is a non-trivial task due to the non-linear key space of Ed25519. The paper BIP32-Ed25519 Hierarchical Deterministic Keys over a Non-linear Keyspace has seen wide adoption in the industry, but has some security issues. In this paper modifications are proposed that addresses those issues, at the cost of reducing the number of levels of child keys.
      \newline
      \newline
    \end{abstract}
  \end{@twocolumnfalse}
]

%%%%%%%%%%%
% Article %
%%%%%%%%%%%

\section{Introduction}

Blockchains rely on users to generate seed phrases that correspond to public/private keypairs for use in signing transactions on-behalf of an address. In the basic case, a single seed phrase corresponds to a single address, requiring users to keep track of many seed phrases should they wish to use separate addresses for separate domains of applications: making and receiving payments, holding NFTs, engaging in a DAO, DeFi, gaming, etc.

\subsection{Hierarchical Deterministic Wallets}

Bitcoin Improvement Proposal 32 (BIP32) was introduced to provide users with a means to deterministically derive many public/private keypairs from a single secret seed, greatly easing the management of seed phrases.

BIP43 introduced the concept of a purpose field, claiming that the original BIP32 standard allowed for too many degrees of freedom and resulted in many wallet softwares which claimed to be BIP32 compatible but were ultimately incompatible with each other. BIP44 took this a step further and defined 5 levels levels in the BIP32 path:

$m / \text{purpose}' / \text{coin\_type}' / \text{account}' / \text{change} / \text{key\_index}$

The apostrophe indicates that the field is hardened, meaning that the child key is derived in a way that prevents the parent public key from being derived from the child public key, or the sibling keys of that child key.

This class of wallets are known as Hierarchical Determinstic Wallets, or HD Wallets for short.

Other derivation schemes exist as well, such as BIP49 for P2SH nested SegWit addresses, and BIP84 for native SegWit addresses.

A complementary set of documents called SatoshiLab Improvement Protocols (SLIP) also exist, maintained by SatoshiLab (the creators of the Trezor hardware wallet) and come up in the context of BIP32 wallets. E.g. SLIP10 or SLIP23. SLIP44 extended BIP44 to assign coin type values for blockchains other than Bitcoin mainnet and testnet, e.g. $\text{283}'$ for Algorand.

\subsection{Elliptic curves and their key spaces}

Bitcoin relies on the Secp256k1 elliptic curve, which has a linear key space. A new public key can be formed by adding $[x]B$ to the public key $[k]B$, where $x$ and $k$ are private keys/scalars and $B$ the base point. This results in a new public key whose private key is $[x] + [k]$ (mod the order of the finite field).

An alternate elliptic curve standard for digital signatures is Ed25519. It is used by a number of blockchains such as Algorand, Cardano, Nano, Monero and more. Polkadot uses a variant called Schnorrkel/Ristretto x25519 ("sr25519") and Trezor offers it alongside other curves.

Ed25519 has a number of advanta\usepackage[skip=10pt]{parskip}ges, such as high speed constant-time implementations and a signing procedure where the nonce is deterministically derived from the data to be signed (mitigating the kind of attack that Sony PS3 suffered).

Unfortunately, private keys have to be derived by hashing the seed and manipulating the bits. Of the 32 byte private key:

- Bits 0, 1, 2:  set to 0, ensuring that the key is a multiple of 8 and avoiding small subgroup attacks.
- Bit 255: set to 1, to ensuring the Montgommery ladder computation used in scalar multiplication takes constant time.
- Bit 254: set to 0, ensuring the key is in the range of the field order ($2^{255} - 19$).

As a result, the key space for Ed25519 is non-linear.

A paper that has seen adoption in the industry is BIP32-Ed25519
Hierarchical Deterministic Keys over a Non-linear Keyspace \cite{BIP32-Ed25519}. It extends the BIP32 standard to work with Ed25519 keys. It has been adopted by Cardano, though conforming additionally with SLIP10 to derive the initial master Ed25519 key and chain code.

While solving the problem (with only minor modifications to BIP32 and working with "after-hash" private keys of Ed25519), in this paper an improvement to their method is proposed that provides a more secure way to derive Ed25519 keys, at the cost of reducing the number of levels of child keys.

\section{Security Review}

The BIP32-Ed25519 by  proposal provides similar features for Ed25519 keys [, which are widely used in <projects>]. However, in the current draft, derived private keys are somewhat closely related, in ways that can significantly degrade security. For example, if even a single "child" private key is compromised, then the security of its "parent" key degrades to only 112--114 bits. In other words, computing the parent key may become more than 64,000 times easier for the attacker. If many sibling private keys are compromised, the parent key's security is degraded even further. And the same applies to further ancestors, up to the root private key.

\section{Suggestions for Improvements}

The purpose of our modifications is to make derived keys less related, so that there is less security degradation if some private keys are compromised. While the BIP32 approach cannot produce completely unrelated Ed25519 keys, our modifications ensure a reduction of only about 3 bits [DEPENDS ON WHICH "g" VALUE YOU CHOSE; IT'S (g-4)/2 BITS] of security in the above scenario. This is done without sacrificing the number of child keys a parent can have, and by retaining a deep enough hierarchy of keys for applications.

\section{Conclusion}

In this paper we have proposed a modification to the BIP32-Ed25519 standard that provides a more secure way to derive Ed25519 keys, at the cost of reducing the number of levels of child keys. We have shown that this modification provides a significant improvement in security, while still providing a sufficient number of child keys for most applications.

%%%%%%%%%%%%%%
% References %
%%%%%%%%%%%%%%

\nocite{*}
\bibliographystyle{plain}
\bibliography{sources}

\end{document}

### 2024-04-30 paper.tex
update

@@
 
-\author{Chris Peikert\cite{Author1}, Yared Afework\cite{Author2}}
+\author{Chris Peikert\cite{Author1}, Yared Efrem Afework\cite{Author2}, Bruno Martins\cite{Author3}}
 
@@
 \titleformat*{\subsection}{\normalsize\bfseries}
-
-%%%%%%%%%%%%%%
-% References %
-%%%%%%%%%%%%%%
-
-% If changing the name of the bib file, change \bibliography{test} at the bottom
-\begin{filecontents}{sources.bib}
-
-@misc{Author1,
-  author       = "Peikert, Chris",
-  howpublished = "\url{mailto:cpeikert@umich.edu}",
-}
-
-@misc{Author2,
-  author       = "Efrem Afework, Yared",
-  howpublished = "\url{mailto:yared.efremafework@algorand.foundation}",
-}
-
-@article{BIP32-Ed25519,
-  author  = "Khovratovich, Dmitry, and Law, Jason",
-  title   = "BIP32-Ed25519 Hierarchical Deterministic Keys over a Non-linear Keyspace",
-  year    = "2017",
-  journal = "IEEE European Symposium on Security and Privacy Workshops (EuroS&PW)",
-  note    = "\url{https://ieeexplore.ieee.org/abstract/document/7966967}",
-}
-
-\end{filecontents}
 
@@
 A new public key can be formed by adding $[x]B$ to the public key $[k]B$, where $x$ and $k$ are private keys/scalars and $B$ the base point.
-This results in a new public key whose private key is $[x] + [k]$ (mod the order of the finite field).
+This results in a new public key whose private key is $[x] + [k] mod q$ for $q$ the order of the field.
 
@@
 
-Ed25519 has a number of advanta\usepackage[skip=10pt]{parskip}ges, such as high speed constant-time implementations and a signing procedure where the nonce is deterministically derived from the data to be signed (mitigating the kind of attack that Sony PS3 suffered).
+Ed25519 has a number of advantages, such as high speed constant-time implementations and a signing procedure where the nonce is deterministically derived from the data to be signed (mitigating the kind of attack that Sony PS3 suffered).
 
@@
 
-- Bits 0, 1, 2:  set to 0, ensuring that the key is a multiple of 8 and avoiding small subgroup attacks.
-- Bit 255: set to 1, to ensuring the Montgommery ladder computation used in scalar multiplication takes constant time.
-- Bit 254: set to 0, ensuring the key is in the range of the field order ($2^{255} - 19$).
+\begin{itemize}
+  \item Bits 0, 1, 2:  set to 0, ensuring that the key is a multiple of 8 and avoiding small subgroup attacks.
+  \item Bit 254: set to 1, to ensuring the Montgommery ladder computation used in scalar multiplication runs in constant time.
+  \item Bit 255: set to 0, ensuring the key is in the range of the field order ($2^{255} - 19$).
+\end{itemize}
 
@@
 
+\section{Implementations}
+
+The suggestions provided in this paper have been implemented as part of the ARC52 Bip32-Ed25519 libraries provided by the Algorand Foundation.
+
+\begin{itemize}
+  \item \href{https://github.com/algorandfoundation/bip32-ed25519-kotlin}{Kotlin}
+  \item \href{https://github.com/algorandfoundation/bip32-ed25519-swift}{Swift}
+  \item \href{https://github.com/algorandfoundation/bip32-ed25519-ts}{TypeScript}
+\end{itemize}
+
 \section{Conclusion}

### 2024-05-02 paper.tex
include sources.bib
28 left bytes

@@
 \usepackage[skip=10pt]{parskip}
+\usepackage{enumitem}
 
@@
 The paper BIP32-Ed25519 Hierarchical Deterministic Keys over a Non-linear Keyspace has seen wide adoption in the industry, but has some security issues.
-In this paper modifications are proposed that addresses those issues, at the cost of reducing the number of levels of child keys.
+In this paper modifications are proposed that addresses those issues, at the cost of reducing the number of possible levels of child keys.
       \newline
@@
 
-$m / \text{purpose}' / \text{coin\_type}' / \text{account}' / \text{change} / \text{key\_index}$
+$m / \text{purpose}' / \text{coin\_type}' / \text{account}' / \text{change} / \text{key}$
 
-The apostrophe indicates that the field is hardened, meaning that the child key is derived in a way that prevents the parent public key from being derived from the child public key, or the sibling keys of that child key.
+\begin{itemize}
+  \item m: master seed
+  \item purpose': derivation path scheme (e.g., 44')
+  \item coin\_type': constant corresponding to the blockchain
+  \item account': user account
+  \item change: 0 for external addresses, 1 for internal addresses
+  \item key: address index
+\end{itemize}
+
+The apostrophe in the paths indicates that the respective field is hardened, meaning that the child key is derived in a way that prevents the parent public key from being derived from the child public key, or the sibling keys of that child key.
+The derivation method is different for hardened keys, requiring the parent private key.
+
+Each level has $2^{32}$ possible values.
+When hardening a level, the first bit is set to 1.
+Thus $[0, 2^{31})$ are reserved for non-hardened keys, and $[2^{31}, 2^{32})$ are reserved for hardened keys.
+
+The terms "account" and "address" are sometimes used interchangeably in casual conversation.
+In the HD wallet context however, a user would create a new account of application and could create a new address index for each transaction
+
+Since the indices and change levels are not hardened but the account level is, it is possible to connect sibling addresses within an account by providing the extended public key produced at the account level, but not connect different accounts with each other should the extended public key at the coin\_type be leaked.
+Similarly, hardening at the coin\_type level means that keys derived for one blockchain cannot be connected to keys derived for another blockchain.
 
@@
 SLIP10 or SLIP23.
-SLIP44 extended BIP44 to assign coin type values for blockchains other than Bitcoin mainnet and testnet, e.g.
+SLIP44 extended BIP44 to assign $\text{coin\_type}$ values for blockchains other than Bitcoin mainnet and testnet, e.g.
 $\text{283}'$ for Algorand.
 
+Note that the schemes listed rely on 5 levels of derivation.
+
 \subsection{Elliptic curves and their key spaces}
+\label{subsec:elliptic_curves_and_their_key_spaces}
 
@@
 A new public key can be formed by adding $[x]B$ to the public key $[k]B$, where $x$ and $k$ are private keys/scalars and $B$ the base point.
-This results in a new public key whose private key is $[x] + [k] mod q$ for $q$ the order of the field.
+This results in a new public key whose private key is $[x] + [k] \text{ mod } q$ for $q$ the order of the field.
 
@@
 
-A paper that has seen adoption in the industry is BIP32-Ed25519
+A paper that has seen large adoption in the industry is BIP32-Ed25519
 Hierarchical Deterministic Keys over a Non-linear Keyspace \cite{BIP32-Ed25519}.
@@
 
-While solving the problem (with only minor modifications to BIP32 and working with "after-hash" private keys of Ed25519), in this paper an improvement to their method is proposed that provides a more secure way to derive Ed25519 keys, at the cost of reducing the number of levels of child keys.
+TODO: add more adopters
+
+While solving the problem (with only minor modifications to BIP32 and working with "after-hash" private keys of Ed25519), certain decisions were made that result in unnecessary security degradation.
+
+In this paper an improvement to their method is proposed that provides a more secure way to derive Ed25519 keys, at the cost of reducing the number of levels of child keys.
 
@@
 
-The BIP32-Ed25519 by  proposal provides similar features for Ed25519 keys [, which are widely used in <projects>].
-However, in the current draft, derived private keys are somewhat closely related, in ways that can significantly degrade security.
-For example, if even a single "child" private key is compromised, then the security of its "parent" key degrades to only 112--114 bits.
+While reviewing the BIP32-Ed25519 paper, two issues were identified, the first minor and the other major.
+
+\subsection{Discarding bad child private or public keys}
+
+While deriving a new private or public child key, it is possible to produce a child that is unsafe to use:
+
+\begin{itemize}
+  \item $k_L$ is divisible by the base order $n$
+  \item $A_i$ is the identity point (0, 1)
+\end{itemize}
+
+The paper states that if these happen the child must be discarded.
+However it does not give instructions on how to proceed with the derivation.
+For example, an implementer could mark the index as forbidden and increment the child index.
+
+Regardless, the likelihood of either happening is highly unlikely.
+
+\subsection{The 28 left bytes of Z }
+
+As was mentioned in \ref{subsec:elliptic_curves_and_their_key_spaces}, there is a need to ensure that bits 0, 1, 2, 3 and 255 are 0 and that bit 254 is 1 for private keys (scalars) in Ed25519.
+
+The paper's method for ensuring this is to produce the child private key respectively public key as follows:
+
+Derive $Z$ from either the parent's public key or parent's extended private key, depending on whether the child is hardened or not:
+
+$  Z \leftarrow 
+  \begin{cases} 
+  F_{c^P}(0x02||A^P||i) & i \in  [0, 2^{31}) \\
+  F_{c^P}(0x00||k^P||i) & i \in [2^{31}, 2^{32})
+  \end{cases}
+  $
+
+Truncate the 32 byte $Z$ to 28 bytes: $Z_L \leftarrow Z_{[:28]}$.
+
+Calculate, depending on choice, the child private key (scalar) or public key (corresponding to an address on the blockchain):
+
+\begin{enumerate}[label={}] 
+  \item $k_L \leftarrow \langle 8[Z_L]  + [k{_L}^P]\rangle$ 
+  \item $A_i \leftarrow A^P + [8Z_L]B$
+\end{enumerate}
+
+The calculations of $k_R$, necessary to produce the extended private key $k = (k_L, k_R)$, and the child chain code $c_i$, necessary to serve as the key for the HMAC-512 function $F$ (and be part of the extended public key $(A_i, c_i)$); have been omitted for brevity.
+$B$ represents the base point of the curve and $i$ a little-endian 4-byte string.
+
+By truncating $Z$ to 28 bytes and then multiplying by 8, the authors ensure that the scalar is a multiple of 8 and that bit 255 is 0. When $k{_L}^P$ is added, it contributes with a 1 on bit 254.
+
+Note that as part of initial conversion from seed to master key $\tilde{k}$, only those $\tilde{k}$ that produce a $k_L$ whose bit 253 is 0 are admitted.
+This prevents overflow when adding, which would affect bit 254.
+
+As a result, the children are guaranteed to be safe in accordance with the requirements of Ed25519, for many levels of derivation. $2^20 = 1048576$ in fact.
+
+Unfortunately, offering this many levels comes at a cost - only 28 bytes of randomization takes place for every level.
+
+If even a single child private key is compromised, then the security of its "parent" key degrades to only 112--114 bits.
 In other words, computing the parent key may become more than 64,000 times easier for the attacker.
@@
 
-\section{Suggestions for Improvements}
+\section{Improving the BIP32-Ed25519 Standard}
 

### 2024-05-03 paper.tex
author
authors
section 2

@@
 
-\author{Chris Peikert\cite{Author1}, Yared Efrem Afework\cite{Author2}, Bruno Martins\cite{Author3}}
+\author{Chris Peikert\cite{Author1}, Algorand Foundation\cite{author2}}
 
@@
 
-As a result, the children are guaranteed to be safe in accordance with the requirements of Ed25519, for many levels of derivation. $2^20 = 1048576$ in fact.
+As a result, the children are guaranteed to be safe in accordance with the requirements of Ed25519, for many levels of derivation.
 
-Unfortunately, offering this many levels comes at a cost - only 28 bytes of randomization takes place for every level.
+To understand how many levels can be safely performed, consider that the root key $a$ can be expressed as follows: $a = 2^{254} + 8b$, where $b < 2^{250}$, due to the bit manipulation and how we generate it.
 
-If even a single child private key is compromised, then the security of its "parent" key degrades to only 112--114 bits.
-In other words, computing the parent key may become more than 64,000 times easier for the attacker.
-If many sibling private keys are compromised, the parent key's security is degraded even further.
-And the same applies to further ancestors, up to the root private key.
+Thus, $a = 2^{254} + 8b < 2^{254} + 2^{253}$.
+
+At the next level, $a_{1} = a + 8[Z_L] < 2^{254} + 2^{253} + 8 \cdot 2^{224} = 2^{254} + 2^{253} + 2^{227}$.
+
+At the next level, $a_{2} = a_{1} + 8[Z_L] < 2^{254} + 2^{253} + 2^{227} + 2^{227} = 2^{254} + 2^{253} + 2^{228}$.
+
+At the next level, $a_{3} = a_{2} + 8[Z_L] < 2^{254} + 2^{253} + 2^{228} + 2^{227}$.
+
+In short, the key at level $j$ is defined by the inequality, $a_j < a + j2^{227} = 2^{254} + 2^{253} + j2^{227}$.
+
+The paper specifies $j \leq 2^{20}$, allowing for a maximum 1048576 levels of derivation, such that $a_{j} \leq 2^{254} + 2^{253} + 2^{247}$ - a form suitable for our purposes.
+
+Unfortunately, offering this many levels comes at a cost: only 28 bytes (224 bits) of randomization takes place for every level.
+
+Consider the scenario where a single child private key is compromised. The security of its parent key would degrade to only $\sqrt{2^{224}}=2^{112}$ due to the Birthday Attack.
+
+If even a single child private key is compromised, then the security of its parent key degrades to only 112--114 bits.
+In other words, computing the parent key may become more than 64,000 times ($\sqrt{2^{32}}=2^{16}$) easier for the attacker as a result of doing away with 4 bytes of entropy.
+
+If many sibling private keys are compromised, the parent key's security is degraded even further. And the same applies to further ancestors, up to the root private key.
 
@@
 The purpose of our modifications is to make derived keys less related, so that there is less security degradation if some private keys are compromised.
+
 While the BIP32 approach cannot produce completely unrelated Ed25519 keys, our modifications ensure a reduction of only about 3 bits [DEPENDS ON WHICH "g" VALUE YOU CHOSE; IT'S (g-4)/2 BITS] of security in the above scenario.
+
 This is done without sacrificing the number of child keys a parent can have, and by retaining a deep enough hierarchy of keys for applications.

### 2024-05-03 paper.tex
conclusion

@@
 \subsection{Hierarchical Deterministic Wallets}
+\label{subsec:hd_wallets}
 
@@
 
-While reviewing the BIP32-Ed25519 paper, two issues were identified, the first minor and the other major.
+While reviewing the BIP32-Ed25519 paper\cite{BIP32-Ed25519}, two issues were identified, the first minor and the other major.
 
@@
 
-Truncate the 32 byte $Z$ to 28 bytes: $Z_L \leftarrow Z_{[:28]}$.
+Truncate the 32 byte $Z$ to 28 bytes: $Z_L \leftarrow Z_{[:224]}$.
 
@@
 
-Consider the scenario where a single child private key is compromised. The security of its parent key would degrade to only $\sqrt{2^{224}}=2^{112}$ due to the Birthday Attack.
+Consider the scenario where a single child private key is compromised.
+The security of its parent key would degrade to around only $\sqrt{2^{224}}=2^{112}$ due to the Birthday Attack, below the 128 bits of security offered by Ed25519.
+This contradicts the author's claims that no loss of security occurs as a result of their method.
 
-If even a single child private key is compromised, then the security of its parent key degrades to only 112--114 bits.
 In other words, computing the parent key may become more than 64,000 times ($\sqrt{2^{32}}=2^{16}$) easier for the attacker as a result of doing away with 4 bytes of entropy.
@@
 
-While the BIP32 approach cannot produce completely unrelated Ed25519 keys, our modifications ensure a reduction of only about 3 bits [DEPENDS ON WHICH "g" VALUE YOU CHOSE; IT'S (g-4)/2 BITS] of security in the above scenario.
+Define $g$, such that $Z_L = Z_{[:256-g]}$.
+The BIP32-Ed25519 paper\cite{BIP32-Ed25519} set $g=32$ to arrive on 28 bytes.
+An Ed25519 scalar has 251 bits available for randomization (255, 254, 2, 1, 0 are fixed), with an additional bit (253) having to be 0 to avoid overflow.
+The level at which we can no longer guarantee 253 will be 0 forms an upper bound.
 
-This is done without sacrificing the number of child keys a parent can have, and by retaining a deep enough hierarchy of keys for applications.
+We thus require, for security and correctness, that the following inequality holds:
+
+\begin{equation*}
+  g \geq d + 6
+\end{equation*}
+
+Where $2^d = D$, the number of levels of derivation. For best results, set the left hand and right hand sides equal.
+
+\begin{table}[h]
+  \centering
+  \begin{tabular}{|c|c|c|}
+  \hline
+  g & d  & D  \\
+  \hline
+  8 & 2 & 4 \\
+  9 & 3 & 8 \\
+  16 & 10 & 1024 \\
+  32 & 20 & 1048576 \\
+  \hline
+  \end{tabular}
+  \caption{Bits to randomize $g$ resulting in levels $D$.}
+  \label{tab:g_and_d}
+  \end{table}
+
+Table \ref{tab:g_and_d} presents a selection of possible values.
+$g=8$ is the most secure and breaks at the byte boundary (= 31 bytes) but results in only 4 levels of derivation - too few to satisfy the derivation schemes mentioned in Section \ref{subsec:hd_wallets} which require 5.
+
+$g=16$ is the next value that breaks at the byte boundary (= 30 bytes) and provides more than enough levels. It randomizes 240 out of the 251 bits available in a Ed25519 scalar.
+
+$g=9$ is the most secure value that also allows for enough derivation levels, allowing for 8 levels of derivation.
+It randomizes 246 out of the 251 bits available in a Ed25519 scalar.
+The only downside is that it does not break at the byte boundary, requiring a little bit of extra care when implementing.
 
@@
 
+In particular, setting $g=9$ allows for the most secure implementation that also satisfies the requirements of the most common derivation schemes.
+
+We have also provided implementations of this modification in several programming languages.
+We hope that this modification will be adopted by the wider community and help to improve the security of Ed25519-based hierarchical deterministic wallets.
+
 %%%%%%%%%%%%%%

### 2024-05-06 paper.tex
2^26 clarification
discarding bad child

@@
 
-The paper states that if these happen the child must be discarded.
-However it does not give instructions on how to proceed with the derivation.
-For example, an implementer could mark the index as forbidden and increment the child index.
+The paper states that if these happen the child must be discarded. However it does not give instructions on how to proceed with the derivation.
+
+For example, an implementer could mark the index as forbidden and simply increment the child index.
+However, depending on the level, should the user share their extended public key to the world and others derive the same faulty child key, the security of the parent key would be degraded.
 
@@
   16 & 10 & 1024 \\
-  32 & 20 & 1048576 \\
+  32 & 26 & 67108864 \\
   \hline
@@
 
+It also becomes apparent that the recommended maximum level of depth in the BIP32-Ed25519 paper, $2^{20}$, is lower than the actual number of levels that can be performed without bit overflows, $2^{26}$.
+Indeed, $a_{j} \leq 2^{254} + 2^{253} + 2^{247} < 2^{254} + 2^{253} + 2^{252}$.
 \section{Implementations}

### 2024-05-06 paper.tex
exponents

@@
 
-The paper specifies $j \leq 2^{20}$, allowing for a maximum 1048576 levels of derivation, such that $a_{j} \leq 2^{254} + 2^{253} + 2^{247}$ - a form suitable for our purposes.
+The paper specifies $j \leq 2^{20}$, allowing for a maximum 1048576 levels of derivation, such that $a_{j} \leq 2^{254} + 2^{253} + 2^{227 + 20 = 247}$ - a form suitable for our purposes.
 
@@
 
-It also becomes apparent that the recommended maximum level of depth in the BIP32-Ed25519 paper, $2^{20}$, is lower than the actual number of levels that can be performed without bit overflows, $2^{26}$.
-Indeed, $a_{j} \leq 2^{254} + 2^{253} + 2^{247} < 2^{254} + 2^{253} + 2^{252}$.
+It also becomes apparent that the recommended maximum level of depth in the BIP32-Ed25519 paper, $2^{20}$, is lower than the actual number of levels that can be performed without bit overflows breaking the scalar's Ed25519-compatibility: $2^{26}$.
+Indeed, $a_{j} \leq 2^{254} + 2^{253} + 2^{247} < 2^{254} + 2^{253} + 2^{227+26=253}$.
 \section{Implementations}

### 2024-05-06 paper.tex
2^255
fixed language, added sources

@@
 
-\author{Chris Peikert\cite{Author1}, Algorand Foundation\cite{author2}}
+% \author{Chris Peikert\cite{ChrisP}}
 
@@
 
-Blockchains rely on users to generate seed phrases that correspond to public/private keypairs for use in signing transactions on-behalf of an address.
-In the basic case, a single seed phrase corresponds to a single address, requiring users to keep track of many seed phrases should they wish to use separate addresses for separate domains of applications: making and receiving payments, holding NFTs, engaging in a DAO, DeFi, gaming, etc.
+Blockchains rely on users to generate seed phrases that correspond to public/private keypairs for use in signing transactions on behalf of an address.
+
+In the basic case, a single seed phrase corresponds to a single address, requiring users to keep track of many seed phrases should they wish to use separate addresses for different application domains: making and receiving payments, holding NFTs, participating in a DAO, DeFi, gaming, etc.
 
@@
 
-Bitcoin Improvement Proposal 32 (BIP32) was introduced to provide users with a means to deterministically derive many public/private keypairs from a single secret seed, greatly easing the management of seed phrases.
+Bitcoin Improvement Proposal 32 (BIP32\cite{BIP32}) was introduced to provide users with a means to deterministically derive many public/private keypairs from a single secret seed, greatly easing the management of seed phrases.
 
-BIP43 introduced the concept of a purpose field, claiming that the original BIP32 standard allowed for too many degrees of freedom and resulted in many wallet softwares which claimed to be BIP32 compatible but were ultimately incompatible with each other.
-BIP44 took this a step further and defined 5 levels levels in the BIP32 path:
+BIP43\cite{BIP43} introduced the concept of a purpose field, claiming that the original BIP32 standard allowed for too many degrees of freedom and resulted in many wallet softwares which claimed to be BIP32 compatible but were ultimately incompatible with each other.
+BIP44\cite{BIP44} took this a step further and defined 5 levels in the BIP32 path:
 
@@
 The terms "account" and "address" are sometimes used interchangeably in casual conversation.
-In the HD wallet context however, a user would create a new account of application and could create a new address index for each transaction
+In the HD wallet context however, an address is the most basic unit, while an account represents a logical grouping of many addresses.
 
-Since the indices and change levels are not hardened but the account level is, it is possible to connect sibling addresses within an account by providing the extended public key produced at the account level, but not connect different accounts with each other should the extended public key at the coin\_type be leaked.
+Change is used to differentiate between addresses used for receiving payments and addresses used for change in transactions, but this is more relevant for UTXO blockchains.
+Account-based blockchains like Algorand or Ethereum do not have this distinction, and can either set it to 0 or reserve it for other use.
+
+Since the indices and change levels are not hardened but the account level is, it is possible to connect sibling addresses within an account by providing the extended public key produced at the account level.
+However, different accounts cannot be connected with each other should the extended public key at the coin\_type level be leaked.
 Similarly, hardening at the coin\_type level means that keys derived for one blockchain cannot be connected to keys derived for another blockchain.
 
-This class of wallets are known as Hierarchical Determinstic Wallets, or HD Wallets for short.
+This class of wallets is known as Hierarchical Deterministic Wallets, or HD Wallets for short.
 
-Other derivation schemes exist as well, such as BIP49 for P2SH nested SegWit addresses, and BIP84 for native SegWit addresses.
+Other derivation schemes exist as well, such as BIP49\cite{BIP49} for P2SH nested SegWit addresses, and BIP84\cite{BIP84} for native SegWit addresses.
+ZIP32\cite{ZIP32}, a variant used by ZCash for their shielded addresses, does away with the change field completely.
 
-A complementary set of documents called SatoshiLab Improvement Protocols (SLIP) also exist, maintained by SatoshiLab (the creators of the Trezor hardware wallet) and come up in the context of BIP32 wallets.
-E.g.
-SLIP10 or SLIP23.
-SLIP44 extended BIP44 to assign $\text{coin\_type}$ values for blockchains other than Bitcoin mainnet and testnet, e.g.
+A complementary set of standards called SatoshiLab Improvement Protocols (SLIP) also exist, maintained by SatoshiLab (the creators of the Trezor hardware wallet) and come up in the context of BIP32 wallets.
+For example, SLIP10\cite{SLIP10} or SLIP23\cite{SLIP23}.
+SLIP44\cite{BIP44} extended BIP44 to assign $\text{coin\_type}$ values for blockchains other than Bitcoin mainnet and testnet, e.g.
 $\text{283}'$ for Algorand.
 
-Note that the schemes listed rely on 5 levels of derivation.
+Note that all of the schemes listed above rely on 5 or fewer levels of derivation.
 
@@
 
-Bitcoin relies on the Secp256k1 elliptic curve, which has a linear key space.
+Bitcoin relies on the Secp256k1\cite{Secp256k1} elliptic curve, which has a linear key space.
 A new public key can be formed by adding $[x]B$ to the public key $[k]B$, where $x$ and $k$ are private keys/scalars and $B$ the base point.
@@
 
-An alternate elliptic curve standard for digital signatures is Ed25519.
-It is used by a number of blockchains such as Algorand, Cardano, Nano, Monero and more.
-Polkadot uses a variant called Schnorrkel/Ristretto x25519 ("sr25519") and Trezor offers it alongside other curves.
+An alternate elliptic curve standard for digital signatures is Ed25519\cite{Ed25519}.
+It is used by a number of blockchains such as Algorand\cite{AlgorandEd25519}, Cardano\cite{CardanoEd25519}, Monero\cite{MoneroEd25519}, Nano\cite{NanoEd25519}, Solana\cite{SolanaEd25519} and more.
+Polkadot\cite{PolkadotEd25519} uses a variant called Schnorrkel/Ristretto x25519 ("sr25519") and Tezos\cite{TezosEd25519} offers it alongside other curves.
 
-Ed25519 has a number of advantages, such as high speed constant-time implementations and a signing procedure where the nonce is deterministically derived from the data to be signed (mitigating the kind of attack that Sony PS3 suffered).
+Ed25519 has a number of advantages, such as high speed constant-time implementations and a signing procedure where the nonce is deterministically derived from the data to be signed (mitigating the kind of attack that Sony PS3 suffered\cite{Ed25519}).
 
-Unfortunately, private keys have to be derived by hashing the seed and manipulating the bits. Of the 32 byte private key:
+Unfortunately, private keys have to be derived by hashing the seed and manipulating the bits. RFC\-8032\cite{RFC8032} specifies that of the 32 byte private key:
 
@@
 
-As a result, the key space for Ed25519 is non-linear.
+By definition the key space for Ed25519 is non-linear.
 
-A paper that has seen large adoption in the industry is BIP32-Ed25519
-Hierarchical Deterministic Keys over a Non-linear Keyspace \cite{BIP32-Ed25519}.
-It extends the BIP32 standard to work with Ed25519 keys.
-It has been adopted by Cardano, though conforming additionally with SLIP10 to derive the initial master Ed25519 key and chain code.
+As a consequence of this fact, the SLIP10\cite{SLIP10} standard decided to not support public key child derivation for Ed25519 keys.
+Only hardened key generation from private parent key to private child key is supported by it.
+This is, as of the time of writing, what Solana supports\cite{SolanaBip32}.
 
-TODO: add more adopters
+In contrast, a paper that has seen recognition in the industry is BIP32-Ed25519
+Hierarchical Deterministic Keys over a Non-linear Keyspace \cite{BIP32-Ed25519}. It extends the BIP32 standard to work with Ed25519 keys with public key child derivation.
 
-While solving the problem (with only minor modifications to BIP32 and working with "after-hash" private keys of Ed25519), certain decisions were made that result in unnecessary security degradation.
+It has been adopted by Cardano, though conforming additionally with SLIP23\cite{SLIP23} to derive the initial master Ed25519 key and chain code.
+
+While solving the problem (with only minor modifications to BIP32 and working with "after-hash" private keys of Ed25519)\cite{BIP32-Ed25519}, certain decisions were made that result in unnecessary security degradation.
 
@@
 
-As was mentioned in \ref{subsec:elliptic_curves_and_their_key_spaces}, there is a need to ensure that bits 0, 1, 2, 3 and 255 are 0 and that bit 254 is 1 for private keys (scalars) in Ed25519.
+As was mentioned in Section \ref{subsec:elliptic_curves_and_their_key_spaces}, there is a need to ensure that bits 0, 1, 2, 3 and 255 are 0 and that bit 254 is 1 for private keys (scalars) in Ed25519.
 
@@
 
-Truncate the 32 byte $Z$ to 28 bytes: $Z_L \leftarrow Z_{[:224]}$.
+Truncate the 32-byte $Z$ to 28 bytes (224 bits): $Z_L \leftarrow Z_{[:224]}$.
 
-Calculate, depending on choice, the child private key (scalar) or public key (corresponding to an address on the blockchain):
+Depending on the choice, calculate the child private key (scalar) or public key (which corresponds to an address on the blockchain):
 
@@
 
-The calculations of $k_R$, necessary to produce the extended private key $k = (k_L, k_R)$, and the child chain code $c_i$, necessary to serve as the key for the HMAC-512 function $F$ (and be part of the extended public key $(A_i, c_i)$); have been omitted for brevity.
+The calculations of $k_R$, which are necessary to produce the extended private key $k = (k_L, k_R)$, and the child chain code $c_i$, which is necessary to serve as the key for the HMAC-512 function $F$ (and be part of the extended public key $(A_i, c_i)$); have been omitted for brevity.
 $B$ represents the base point of the curve and $i$ a little-endian 4-byte string.
 
-By truncating $Z$ to 28 bytes and then multiplying by 8, the authors ensure that the scalar is a multiple of 8 and that bit 255 is 0. When $k{_L}^P$ is added, it contributes with a 1 on bit 254.
+By truncating $Z$ to 28 bytes and then multiplying by 8, the authors ensure that the scalar is a multiple of 8 and that bit 255 is 0. When $k{_L}^P$ is added, it contributes a 1 on bit 254.
 
-Note that as part of initial conversion from seed to master key $\tilde{k}$, only those $\tilde{k}$ that produce a $k_L$ whose bit 253 is 0 are admitted.
+Note that as part of the initial conversion from seed to master key $\tilde{k}$, only those $\tilde{k}$ that produce a $k_L$ with bit 253 set to 0 are admitted.
 This prevents overflow when adding, which would affect bit 254.
@@
 
-To understand how many levels can be safely performed, consider that the root key $a$ can be expressed as follows: $a = 2^{254} + 8b$, where $b < 2^{250}$, due to the bit manipulation and how we generate it.
+To understand how many levels can be safely performed, consider that the root key $a$ can be expressed as follows: $a = 2^{254} + 8b$, where $b < 2^{250}$, due to the bit manipulation and the way it is generated.
 
@@
 Table \ref{tab:g_and_d} presents a selection of possible values.
-$g=8$ is the most secure and breaks at the byte boundary (= 31 bytes) but results in only 4 levels of derivation - too few to satisfy the derivation schemes mentioned in Section \ref{subsec:hd_wallets} which require 5.
+$g=8$ is the most secure and breaks at the byte boundary (256 bits - 8 bits = 31 bytes) but it results in only 4 levels of derivation - too few to satisfy the derivation schemes mentioned in Section \ref{subsec:hd_wallets} which require 5.
 
-$g=16$ is the next value that breaks at the byte boundary (= 30 bytes) and provides more than enough levels. It randomizes 240 out of the 251 bits available in a Ed25519 scalar.
+$g=16$ is the next value that breaks at the byte boundary (equal to 30 bytes) and provides more than enough levels. It randomizes 240 out of the 251 bits available in an Ed25519 scalar.
 
@@
 It randomizes 246 out of the 251 bits available in a Ed25519 scalar.
-The only downside is that it does not break at the byte boundary, requiring a little bit of extra care when implementing.
+The only downside is that it does not break at the byte boundary, requiring a bit of extra care when implementing.
 
 It also becomes apparent that the recommended maximum level of depth in the BIP32-Ed25519 paper, $2^{20}$, is lower than the actual number of levels that can be performed without bit overflows breaking the scalar's Ed25519-compatibility: $2^{26}$.
-Indeed, $a_{j} \leq 2^{254} + 2^{253} + 2^{247} < 2^{254} + 2^{253} + 2^{227+26=253}$.
+Indeed, $a_{j} \leq 2^{254} + 2^{253} + 2^{247} < 2^{254} + 2^{253} + 2^{227+26=253} = 2^{255}$.
 \section{Implementations}
 
-The suggestions provided in this paper have been implemented as part of the ARC52 Bip32-Ed25519 libraries provided by the Algorand Foundation.
+The suggestions provided in this paper have been implemented as part of the ARC52 Bip32-Ed25519 libraries provided by the Algorand Foundation \cite{AF}.
 
@@
   \item \href{https://github.com/algorandfoundation/bip32-ed25519-swift}{Swift}
-  \item \href{https://github.com/algorandfoundation/bip32-ed25519-ts}{TypeScript}
+  \item \href{https://github.com/ehanoc/ARCs/tree/wallet-api-context/assets/arc-0052}{TypeScript}
 \end{itemize}
@@
 
-In this paper we have proposed a modification to the BIP32-Ed25519 standard that provides a more secure way to derive Ed25519 keys, at the cost of reducing the number of levels of child keys.
+In this paper a modification to the BIP32-Ed25519 standard has been presented that provides a more secure way to derive Ed25519 keys, at the cost of reducing the number of levels of child keys.
 We have shown that this modification provides a significant improvement in security, while still providing a sufficient number of child keys for most applications.

### 2024-05-06 paper.tex
dashes
adds analysis

@@
 
-\begin{itemize}
+\begin{itemize}[label=\textendash, itemsep=-0.5em]
   \item m: master seed
@@
 
-\begin{itemize}
+\begin{itemize}[label=\textendash, itemsep=-0.5em]
   \item Bits 0, 1, 2:  set to 0, ensuring that the key is a multiple of 8 and avoiding small subgroup attacks.
@@
 
-\begin{itemize}
+\begin{itemize}[label=\textendash, itemsep=-0.5em]
   \item $k_L$ is divisible by the base order $n$
@@
 
-\subsection{The 28 left bytes of Z }
+\subsection{The 28 left bytes of Z}
+\label{subsec:the_28_left_bytes_of_z}
 
@@
 
-In short, the key at level $j$ is defined by the inequality, $a_j < a + j2^{227} = 2^{254} + 2^{253} + j2^{227}$.
+In short, the key at level $D$ depth (called $j$ in the paper) is defined by the inequality, $a_D < a + D\cdot2^{227} = 2^{254} + 2^{253} + D\cdot2^{227}$.
 
-The paper specifies $j \leq 2^{20}$, allowing for a maximum 1048576 levels of derivation, such that $a_{j} \leq 2^{254} + 2^{253} + 2^{227 + 20 = 247}$ - a form suitable for our purposes.
+The paper specifies $D \leq 2^{20}$, allowing for a maximum 1048576 levels of derivation, such that $a_{D} \leq 2^{254} + 2^{253} + 2^{227 + 20 = 247}$ - a form suitable for our purposes.
 
@@
 
-Define $g$, such that $Z_L = Z_{[:256-g]}$.
-The BIP32-Ed25519 paper\cite{BIP32-Ed25519} set $g=32$ to arrive on 28 bytes.
+Define $g$, such that $Z_L = Z_{[:256-g]}$. The BIP32-Ed25519 paper\cite{BIP32-Ed25519} set $g=32$ to arrive on 28 bytes.
+
 An Ed25519 scalar has 251 bits available for randomization (255, 254, 2, 1, 0 are fixed), with an additional bit (253) having to be 0 to avoid overflow.
@@
 
-We thus require, for security and correctness, that the following inequality holds:
+Recall from Section \ref{subsec:the_28_left_bytes_of_z} that each $Z_L < 2^{256 - g}$ and is multiplied by 8.
+Root $a$ must also be less than $2^{254} + 2^{253}$.
+Define $d = log_2(D)$.
+The previous paragraph can thus be expressed as follows:
 
-\begin{equation*}
-  g \geq d + 6
-\end{equation*}
+\begin{align*}
+  a_D &< a + 8\cdot D\cdot2^{256 - g} \\
+  a_D &< a + D\cdot2^{259 - g} \\
+  a_D &< a + 2^{259 + d - g}\\
+  a_D < 2^{254} +& 2^{253} + 2^{259 + d - g} < 2 \cdot 2^{254}
+\end{align*}
 
-Where $2^d = D$, the number of levels of derivation. For best results, set the left hand and right hand sides equal.
+The relationship between $d$ and $g$ can be expressed as follows:
+
+\begin{align*}
+  253 &\geq 259 + d - g \\
+  &\Downarrow \\
+  g &\geq d + 6
+\end{align*}
+
+ For best results, set the left hand and right hand sides equal.
 
@@
 It also becomes apparent that the recommended maximum level of depth in the BIP32-Ed25519 paper, $2^{20}$, is lower than the actual number of levels that can be performed without bit overflows breaking the scalar's Ed25519-compatibility: $2^{26}$.
-Indeed, $a_{j} \leq 2^{254} + 2^{253} + 2^{247} < 2^{254} + 2^{253} + 2^{227+26=253} = 2^{255}$.
+
+Indeed, for $D=2^{20}$, $a_{D} \leq 2^{254} + 2^{253} + 2^{247} < 2^{254} + 2^{253} + 2^{227+26=253} = 2^{255}$.
+
 \section{Implementations}

### 2024-05-06 paper.tex
whitespace
wording
clarity
with yareds correction

@@
 The previous paragraph can thus be expressed as follows:
-
 \begin{align*}
@@
   a_D &< a + 2^{259 + d - g}\\
-  a_D < 2^{254} +& 2^{253} + 2^{259 + d - g} < 2 \cdot 2^{254}
+  a_D < 2^{254} +& 2^{253} + 2^{259 + d - g} < 2^{254} + 2^{254}
+\end{align*}
+The relationship between $d$ and $g$ can be stated as:
+\begin{align*}
+  253 &> 259 + d - g \\
+  &\Downarrow \\
+  g &> d + 6
 \end{align*}
 
-The relationship between $d$ and $g$ can be expressed as follows:
+ $g = d + 6$ defines the first level at which the 3rd highest bit overflows into the 2nd highest bit, which is set to 1 already and would overflow into the highest bit.
+$d$ must necessarily be $d < g - 6$.
 
-\begin{align*}
-  253 &\geq 259 + d - g \\
-  &\Downarrow \\
-  g &\geq d + 6
-\end{align*}
+ As a result, the number of safe levels $S$ of derivation that can be performed is one level fewer than $D = 2^{d} < 2^{g - 6}$:
 
- For best results, set the left hand and right hand sides equal.
+ \begin{equation*}
+  S = D - 1 = 2^{d} - 1 \leq 2^{g - 6}
+  \end{equation*}
 
@@
   \centering
-  \begin{tabular}{|c|c|c|}
+  \begin{tabular}{|c|c|c|c|}
   \hline
-  g & d  & D  \\
+  g & d  & D & S  \\
   \hline
-  8 & 2 & 4 \\
-  9 & 3 & 8 \\
-  16 & 10 & 1024 \\
-  32 & 26 & 67108864 \\
+  8 & 2 & 4 & 3 \\
+  9 & 3 & 8 & 7 \\
+  16 & 10 & 1024 & 1023 \\
+  32 & 26 & 67108864 & 67108863  \\
   \hline
   \end{tabular}
-  \caption{Bits to randomize $g$ resulting in levels $D$.}
+  \caption{Bits to randomize $g$ resulting in levels $D$. $S$ is the number of safe levels.}
   \label{tab:g_and_d}
@@
 Table \ref{tab:g_and_d} presents a selection of possible values.
-$g=8$ is the most secure and breaks at the byte boundary (256 bits - 8 bits = 31 bytes) but it results in only 4 levels of derivation - too few to satisfy the derivation schemes mentioned in Section \ref{subsec:hd_wallets} which require 5.
+$g=8$ is the most secure and breaks at the byte boundary (256 bits - 8 bits = 31 bytes) but it results in only 3 levels of derivation - too few to satisfy the derivation schemes mentioned in Section \ref{subsec:hd_wallets} which require 5.
 
@@
 
-$g=9$ is the most secure value that also allows for enough derivation levels, allowing for 8 levels of derivation.
+$g=9$ is the most secure value that also allows for enough derivation levels, allowing for 7 levels of derivation.
 It randomizes 246 out of the 251 bits available in a Ed25519 scalar.
@@
 
-It also becomes apparent that the recommended maximum level of depth in the BIP32-Ed25519 paper, $2^{20}$, is lower than the actual number of levels that can be performed without bit overflows breaking the scalar's Ed25519-compatibility: $2^{26}$.
+It also becomes apparent that the recommended maximum level of depth in the BIP32-Ed25519 paper, $2^{20}$, is lower than the actual number of levels that can be performed without bit overflows breaking the scalar's Ed25519-compatibility: $2^{26} - 1$.
 

### 2024-05-08 paper.tex
yareds corrections - remove unnecessary ineq
rephrase

@@
       Implementing BIP32-style HD wallets for Ed25519 keys is a non-trivial task due to the non-linear key space of Ed25519.
-The paper BIP32-Ed25519 Hierarchical Deterministic Keys over a Non-linear Keyspace has seen wide adoption in the industry, but has some security issues.
+The paper BIP32-Ed25519 Hierarchical Deterministic Keys over a Non-linear Keyspace has seen adoption in the industry, but has some security issues.
 In this paper modifications are proposed that addresses those issues, at the cost of reducing the number of possible levels of child keys.
@@
 
-The apostrophe in the paths indicates that the respective field is hardened, meaning that the child key is derived in a way that prevents the parent public key from being derived from the child public key, or the sibling keys of that child key.
+The apostrophe in the paths indicates that the respective field is hardened, meaning that the child key is derived in a way that prevents the child public key from being derived publicly from the parent public key, or the sibling keys of that child key.
 The derivation method is different for hardened keys, requiring the parent private key.
@@
  \begin{equation*}
-  S = D - 1 = 2^{d} - 1 \leq 2^{g - 6}
+  S = D - 1 = 2^{d} - 1
   \end{equation*}

### 2024-05-16 paper.tex
fix quotation marks

@@
 
-The terms "account" and "address" are sometimes used interchangeably in casual conversation.
+The terms ``account'' and ``address'' are sometimes used interchangeably in casual conversation.
 In the HD wallet context however, an address is the most basic unit, while an account represents a logical grouping of many addresses.
@@
 It is used by a number of blockchains such as Algorand\cite{AlgorandEd25519}, Cardano\cite{CardanoEd25519}, Monero\cite{MoneroEd25519}, Nano\cite{NanoEd25519}, Solana\cite{SolanaEd25519} and more.
-Polkadot\cite{PolkadotEd25519} uses a variant called Schnorrkel/Ristretto x25519 ("sr25519") and Tezos\cite{TezosEd25519} offers it alongside other curves.
+Polkadot\cite{PolkadotEd25519} uses a variant called Schnorrkel/Ristretto x25519 (``sr25519'') and Tezos\cite{TezosEd25519} offers it alongside other curves.
 
@@
 
-While solving the problem (with only minor modifications to BIP32 and working with "after-hash" private keys of Ed25519)\cite{BIP32-Ed25519}, certain decisions were made that result in unnecessary security degradation.
+While solving the problem (with only minor modifications to BIP32 and working with ``after-hash'' private keys of Ed25519)\cite{BIP32-Ed25519}, certain decisions were made that result in unnecessary security degradation.
 
@@
 \end{document}
+
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: t
+%%% End:

### 2024-05-16 paper.tex
use standard bracketed refs for citations

@@
 \usepackage{xurl}
-\usepackage[super,comma,sort&compress]{natbib}
+\usepackage[sort&compress]{natbib}
 \usepackage{abstract}
@@
 
-Bitcoin Improvement Proposal 32 (BIP32\cite{BIP32}) was introduced to provide users with a means to deterministically derive many public/private keypairs from a single secret seed, greatly easing the management of seed phrases.
+Bitcoin Improvement Proposal 32 (BIP32~\cite{BIP32}) was introduced to provide users with a means to deterministically derive many public/private keypairs from a single secret seed, greatly easing the management of seed phrases.
 
-BIP43\cite{BIP43} introduced the concept of a purpose field, claiming that the original BIP32 standard allowed for too many degrees of freedom and resulted in many wallet softwares which claimed to be BIP32 compatible but were ultimately incompatible with each other.
-BIP44\cite{BIP44} took this a step further and defined 5 levels in the BIP32 path:
+BIP43~\cite{BIP43} introduced the concept of a purpose field, claiming that the original BIP32 standard allowed for too many degrees of freedom and resulted in many wallet softwares which claimed to be BIP32 compatible but were ultimately incompatible with each other.
+BIP44~\cite{BIP44} took this a step further and defined 5 levels in the BIP32 path:
 
@@
 
-Other derivation schemes exist as well, such as BIP49\cite{BIP49} for P2SH nested SegWit addresses, and BIP84\cite{BIP84} for native SegWit addresses.
-ZIP32\cite{ZIP32}, a variant used by ZCash for their shielded addresses, does away with the change field completely.
+Other derivation schemes exist as well, such as BIP49~\cite{BIP49} for P2SH nested SegWit addresses, and BIP84~\cite{BIP84} for native SegWit addresses.
+ZIP32~\cite{ZIP32}, a variant used by ZCash for their shielded addresses, does away with the change field completely.
 
 A complementary set of standards called SatoshiLab Improvement Protocols (SLIP) also exist, maintained by SatoshiLab (the creators of the Trezor hardware wallet) and come up in the context of BIP32 wallets.
-For example, SLIP10\cite{SLIP10} or SLIP23\cite{SLIP23}.
-SLIP44\cite{BIP44} extended BIP44 to assign $\text{coin\_type}$ values for blockchains other than Bitcoin mainnet and testnet, e.g.
+For example, SLIP10~\cite{SLIP10} or SLIP23~\cite{SLIP23}.
+SLIP44~\cite{BIP44} extended BIP44 to assign $\text{coin\_type}$ values for blockchains other than Bitcoin mainnet and testnet, e.g.
 $\text{283}'$ for Algorand.
@@
 
-Bitcoin relies on the Secp256k1\cite{Secp256k1} elliptic curve, which has a linear key space.
+Bitcoin relies on the Secp256k1~\cite{Secp256k1} elliptic curve, which has a linear key space.
 A new public key can be formed by adding $[x]B$ to the public key $[k]B$, where $x$ and $k$ are private keys/scalars and $B$ the base point.
@@
 
-An alternate elliptic curve standard for digital signatures is Ed25519\cite{Ed25519}.
-It is used by a number of blockchains such as Algorand\cite{AlgorandEd25519}, Cardano\cite{CardanoEd25519}, Monero\cite{MoneroEd25519}, Nano\cite{NanoEd25519}, Solana\cite{SolanaEd25519} and more.
-Polkadot\cite{PolkadotEd25519} uses a variant called Schnorrkel/Ristretto x25519 (``sr25519'') and Tezos\cite{TezosEd25519} offers it alongside other curves.
+An alternate elliptic curve standard for digital signatures is Ed25519~\cite{Ed25519}.
+It is used by a number of blockchains such as Algorand~\cite{AlgorandEd25519}, Cardano~\cite{CardanoEd25519}, Monero~\cite{MoneroEd25519}, Nano~\cite{NanoEd25519}, Solana~\cite{SolanaEd25519} and more.
+Polkadot~\cite{PolkadotEd25519} uses a variant called Schnorrkel/Ristretto x25519 (``sr25519'') and Tezos~\cite{TezosEd25519} offers it alongside other curves.
 
-Ed25519 has a number of advantages, such as high speed constant-time implementations and a signing procedure where the nonce is deterministically derived from the data to be signed (mitigating the kind of attack that Sony PS3 suffered\cite{Ed25519}).
+Ed25519 has a number of advantages, such as high speed constant-time implementations and a signing procedure where the nonce is deterministically derived from the data to be signed (mitigating the kind of attack that Sony PS3 suffered~\cite{Ed25519}).
 
-Unfortunately, private keys have to be derived by hashing the seed and manipulating the bits. RFC\-8032\cite{RFC8032} specifies that of the 32 byte private key:
+Unfortunately, private keys have to be derived by hashing the seed and manipulating the bits. RFC\-8032~\cite{RFC8032} specifies that of the 32 byte private key:
 
@@
 
-As a consequence of this fact, the SLIP10\cite{SLIP10} standard decided to not support public key child derivation for Ed25519 keys.
+As a consequence of this fact, the SLIP10~\cite{SLIP10} standard decided to not support public key child derivation for Ed25519 keys.
 Only hardened key generation from private parent key to private child key is supported by it.
-This is, as of the time of writing, what Solana supports\cite{SolanaBip32}.
+This is, as of the time of writing, what Solana supports~\cite{SolanaBip32}.
 
 In contrast, a paper that has seen recognition in the industry is BIP32-Ed25519
-Hierarchical Deterministic Keys over a Non-linear Keyspace \cite{BIP32-Ed25519}. It extends the BIP32 standard to work with Ed25519 keys with public key child derivation.
+Hierarchical Deterministic Keys over a Non-linear Keyspace~\cite{BIP32-Ed25519}. It extends the BIP32 standard to work with Ed25519 keys with public key child derivation.
 
-It has been adopted by Cardano, though conforming additionally with SLIP23\cite{SLIP23} to derive the initial master Ed25519 key and chain code.
+It has been adopted by Cardano, though conforming additionally with SLIP23~\cite{SLIP23} to derive the initial master Ed25519 key and chain code.
 
-While solving the problem (with only minor modifications to BIP32 and working with ``after-hash'' private keys of Ed25519)\cite{BIP32-Ed25519}, certain decisions were made that result in unnecessary security degradation.
+While solving the problem (with only minor modifications to BIP32 and working with ``after-hash'' private keys of Ed25519)~\cite{BIP32-Ed25519}, certain decisions were made that result in unnecessary security degradation.
 
@@
 
-While reviewing the BIP32-Ed25519 paper\cite{BIP32-Ed25519}, two issues were identified, the first minor and the other major.
+While reviewing the BIP32-Ed25519 paper~\cite{BIP32-Ed25519}, two issues were identified, the first minor and the other major.
 
@@
 
-Define $g$, such that $Z_L = Z_{[:256-g]}$. The BIP32-Ed25519 paper\cite{BIP32-Ed25519} set $g=32$ to arrive on 28 bytes.
+Define $g$, such that $Z_L = Z_{[:256-g]}$. The BIP32-Ed25519 paper~\cite{BIP32-Ed25519} set $g=32$ to arrive on 28 bytes.
 
@@
 
-The suggestions provided in this paper have been implemented as part of the ARC52 Bip32-Ed25519 libraries provided by the Algorand Foundation \cite{AF}.
+The suggestions provided in this paper have been implemented as part of the ARC52 Bip32-Ed25519 libraries provided by the Algorand Foundation~\cite{AF}.
 

### 2024-05-16 paper.tex
use cleveref for internal refs

@@
 \hypersetup{colorlinks=true, urlcolor=blue, linkcolor=blue, citecolor=blue}
+\usepackage[capitalize,nameinlink,noabbrev]{cleveref} % must load after hyperref
 
@@
 
-As was mentioned in Section \ref{subsec:elliptic_curves_and_their_key_spaces}, there is a need to ensure that bits 0, 1, 2, 3 and 255 are 0 and that bit 254 is 1 for private keys (scalars) in Ed25519.
+As was mentioned in \cref{subsec:elliptic_curves_and_their_key_spaces}, there is a need to ensure that bits 0, 1, 2, 3 and 255 are 0 and that bit 254 is 1 for private keys (scalars) in Ed25519.
 
@@
 
-Recall from Section \ref{subsec:the_28_left_bytes_of_z} that each $Z_L < 2^{256 - g}$ and is multiplied by 8.
+Recall from \cref{subsec:the_28_left_bytes_of_z} that each $Z_L < 2^{256 - g}$ and is multiplied by 8.
 Root $a$ must also be less than $2^{254} + 2^{253}$.
@@
 
-Table \ref{tab:g_and_d} presents a selection of possible values.
-$g=8$ is the most secure and breaks at the byte boundary (256 bits - 8 bits = 31 bytes) but it results in only 3 levels of derivation - too few to satisfy the derivation schemes mentioned in Section \ref{subsec:hd_wallets} which require 5.
+\Cref{tab:g_and_d} presents a selection of possible values.
+$g=8$ is the most secure and breaks at the byte boundary (256 bits - 8 bits = 31 bytes) but it results in only 3 levels of derivation - too few to satisfy the derivation schemes mentioned in \cref{subsec:hd_wallets}, which require 5.
 

### 2024-05-23 paper.tex
use microtype
fix description of private key (scalar) clamping; staying < 2^255-19 is not a relevant requirement

@@
 \documentclass[12pt, a4paper, twocolumn]{article}
+\usepackage{microtype}
 \usepackage{xurl}
@@
 
-Unfortunately, private keys have to be derived by hashing the seed and manipulating the bits. RFC\-8032~\cite{RFC8032} specifies that of the 32 byte private key:
-
+Unfortunately, private keys have to be derived by hashing the seed and manipulating the bits.
+RFC\-8032~\cite{RFC8032} specifies that of the 32-byte ``expanded'' private key:
 \begin{itemize}[label=\textendash, itemsep=-0.5em]
-  \item Bits 0, 1, 2:  set to 0, ensuring that the key is a multiple of 8 and avoiding small subgroup attacks.
-  \item Bit 254: set to 1, to ensuring the Montgommery ladder computation used in scalar multiplication runs in constant time.
-  \item Bit 255: set to 0, ensuring the key is in the range of the field order ($2^{255} - 19$).
+\item Bits 0, 1, 2 are set to 0, ensuring that the key is a multiple of 8, thus avoiding small-subgroup attacks.
+\item Bit 255 is set to 0, to ensure that the key is less than the order of the elliptic-curve group.
+\item Bit 254 is to 1, to ensure that the Montgomery ladder computation used in scalar multiplication runs in constant time.
 \end{itemize}
-
-By definition the key space for Ed25519 is non-linear.
+So, by definition, the key space for Ed25519 is non-linear.
 

### 2024-05-23 paper.tex
clean up description and typesetting of derivation and fields

@@
 BIP44~\cite{BIP44} took this a step further and defined 5 levels in the BIP32 path:
-
-$m / \text{purpose}' / \text{coin\_type}' / \text{account}' / \text{change} / \text{key}$
+\begin{verbatim}
+m / purpose' / coin_type' /
+    account' / change / key
+\end{verbatim}
 
 \begin{itemize}[label=\textendash, itemsep=-0.5em]
-  \item m: master seed
-  \item purpose': derivation path scheme (e.g., 44')
-  \item coin\_type': constant corresponding to the blockchain
-  \item account': user account
-  \item change: 0 for external addresses, 1 for internal addresses
-  \item key: address index
+  \item \verb=m=: master seed
+  \item \verb=purpose'=: derivation path scheme (e.g., 44')
+  \item \verb=coin_type'=: constant corresponding to the blockchain
+  \item \verb=account'=: user account
+  \item \verb=change=: 0 for external addresses, 1 for internal addresses
+  \item \verb=key=: address index
 \end{itemize}
 
-The apostrophe in the paths indicates that the respective field is hardened, meaning that the child key is derived in a way that prevents the child public key from being derived publicly from the parent public key, or the sibling keys of that child key.
+The apostrophe in some fields indicates that the field is \emph{hardened}, meaning that the child key is derived in a way that prevents the child public key from being derived publicly from the parent public key, or the sibling keys of that child key.
 The derivation method is different for hardened keys, requiring the parent private key.
@@
 Each level has $2^{32}$ possible values.
-When hardening a level, the first bit is set to 1.
+When hardening a level, the most-significant bit is set to 1.
 Thus $[0, 2^{31})$ are reserved for non-hardened keys, and $[2^{31}, 2^{32})$ are reserved for hardened keys.
@@
 The terms ``account'' and ``address'' are sometimes used interchangeably in casual conversation.
-In the HD wallet context however, an address is the most basic unit, while an account represents a logical grouping of many addresses.
+However, in the HD wallet context, an address is the most basic unit, while an account represents a logical grouping of many addresses.
 
-Change is used to differentiate between addresses used for receiving payments and addresses used for change in transactions, but this is more relevant for UTXO blockchains.
+The \verb=change= field is used to differentiate between addresses used for receiving payments and addresses used for change in transactions, but this is more relevant for UTXO blockchains.
 Account-based blockchains like Algorand or Ethereum do not have this distinction, and can either set it to 0 or reserve it for other use.
 
-Since the indices and change levels are not hardened but the account level is, it is possible to connect sibling addresses within an account by providing the extended public key produced at the account level.
-However, different accounts cannot be connected with each other should the extended public key at the coin\_type level be leaked.
-Similarly, hardening at the coin\_type level means that keys derived for one blockchain cannot be connected to keys derived for another blockchain.
+Since the indices and change levels are not hardened, but the account level is, it is possible to connect sibling addresses within an account by providing the extended public key produced at the account level.
+However, different accounts cannot be connected with each other should the extended public key at the \verb=coin_type= level be leaked.
+Similarly, hardening at the \verb=coin_type= level means that keys derived for one blockchain cannot be connected to keys derived for another blockchain.
 
@@
 For example, SLIP10~\cite{SLIP10} or SLIP23~\cite{SLIP23}.
-SLIP44~\cite{BIP44} extended BIP44 to assign $\text{coin\_type}$ values for blockchains other than Bitcoin mainnet and testnet, e.g.
-$\text{283}'$ for Algorand.
+SLIP44~\cite{BIP44} extended BIP44 to assign \verb=coin_type= values for blockchains other than Bitcoin mainnet and testnet, e.g. \verb=283'= for Algorand.
 
-Note that all of the schemes listed above rely on 5 or fewer levels of derivation.
+Note that all of the schemes listed above rely on five or fewer levels of derivation.
 

### 2024-05-23 paper.tex
clean up intro description of Secp and Ed25519

@@
 Bitcoin relies on the Secp256k1~\cite{Secp256k1} elliptic curve, which has a linear key space.
-A new public key can be formed by adding $[x]B$ to the public key $[k]B$, where $x$ and $k$ are private keys/scalars and $B$ the base point.
-This results in a new public key whose private key is $[x] + [k] \text{ mod } q$ for $q$ the order of the field.
+A new public key can be formed by adding $[x]B$ to the public key $[k]B$, where $x$ and $k$ are private keys (scalars), and $B$ is the base point.
+This results in a new public key whose private key is $[x] + [k] \bmod q$, where $q$ is the order of the field.
 
-An alternate elliptic curve standard for digital signatures is Ed25519~\cite{Ed25519}.
-It is used by a number of blockchains such as Algorand~\cite{AlgorandEd25519}, Cardano~\cite{CardanoEd25519}, Monero~\cite{MoneroEd25519}, Nano~\cite{NanoEd25519}, Solana~\cite{SolanaEd25519} and more.
+An alternative elliptic-curve standard for digital signatures is Ed25519~\cite{Ed25519}.
+It is used by a number of blockchains such as Algorand~\cite{AlgorandEd25519}, Cardano~\cite{CardanoEd25519}, Monero~\cite{MoneroEd25519}, Nano~\cite{NanoEd25519}, Solana~\cite{SolanaEd25519}, and more.
 Polkadot~\cite{PolkadotEd25519} uses a variant called Schnorrkel/Ristretto x25519 (``sr25519'') and Tezos~\cite{TezosEd25519} offers it alongside other curves.
 
-Ed25519 has a number of advantages, such as high speed constant-time implementations and a signing procedure where the nonce is deterministically derived from the data to be signed (mitigating the kind of attack that Sony PS3 suffered~\cite{Ed25519}).
+Ed25519 has a number of advantages, such as high-speed constant-time implementations, and a signing procedure where the nonce is deterministically derived from the data to be signed (mitigating the kind of attack that Sony PS3 suffered).
 

### 2024-05-23 paper.tex
clean up and correct "clamping" description

@@
 
-Unfortunately, private keys have to be derived by hashing the seed and manipulating the bits.
-RFC\-8032~\cite{RFC8032} specifies that of the 32-byte ``expanded'' private key:
+Unfortunately, private keys in Ed25519 have to be derived by hashing a private ``seed'' and manipulating the bits.
+RFC\-8032~\cite{RFC8032} specifies that in the 32-byte private key:
 \begin{itemize}[label=\textendash, itemsep=-0.5em]
@@
 \item Bit 255 is set to 0, to ensure that the key is less than the order of the elliptic-curve group.
-\item Bit 254 is to 1, to ensure that the Montgomery ladder computation used in scalar multiplication runs in constant time.
+\item Bit 254 is set to 1, to ensure that the Montgomery ladder computation used in scalar multiplication runs in constant time.
 \end{itemize}
-So, by definition, the key space for Ed25519 is non-linear.
+So, by definition, the key space for Ed25519 is non-linear relative to the group order.
 

### 2024-05-23 paper.tex
clean up intro to BIP32-Ed25519

@@
 
-As a consequence of this fact, the SLIP10~\cite{SLIP10} standard decided to not support public key child derivation for Ed25519 keys.
+As a consequence, the SLIP10~\cite{SLIP10} standard decided to not support public key child derivation for Ed25519 keys.
 Only hardened key generation from private parent key to private child key is supported by it.
@@
 
-In contrast, a paper that has seen recognition in the industry is BIP32-Ed25519
-Hierarchical Deterministic Keys over a Non-linear Keyspace~\cite{BIP32-Ed25519}. It extends the BIP32 standard to work with Ed25519 keys with public key child derivation.
-
+In contrast, a paper that has seen recognition in the industry is ``BIP32-Ed25519 Hierarchical Deterministic Keys over a Non-linear Keyspace''~\cite{BIP32-Ed25519}.
+It extends the BIP32 standard to work with Ed25519 keys with public-key child derivation.
 It has been adopted by Cardano, though conforming additionally with SLIP23~\cite{SLIP23} to derive the initial master Ed25519 key and chain code.
 
-While solving the problem (with only minor modifications to BIP32 and working with ``after-hash'' private keys of Ed25519)~\cite{BIP32-Ed25519}, certain decisions were made that result in unnecessary security degradation.
+While solving the problem (with only minor modifications to BIP32 and working with ``after-hash'' private keys of Ed25519)~\cite{BIP32-Ed25519}, it has certain design aspects that result in unnecessary security degradation.
 

### 2024-05-23 paper.tex
clean up description of minor issue

@@
 
-While deriving a new private or public child key, it is possible to produce a child that is unsafe to use:
-
+While deriving a new private or public child key, it is possible to produce a child that is unsafe to use, in the following two cases:
 \begin{itemize}[label=\textendash, itemsep=-0.5em]
-  \item $k_L$ is divisible by the base order $n$
-  \item $A_i$ is the identity point (0, 1)
+\item $k_L$ is divisible by the order of the base point,
+\item $A_i$ is the identity point.
 \end{itemize}
 
-The paper states that if these happen the child must be discarded. However it does not give instructions on how to proceed with the derivation.
+The paper states that if either of these happen, the child must be discarded.
+However, it does not give instructions on how to proceed with the derivation.
+For example, an implementer could mark the index as forbidden and simply increment the child index.
+However, depending on the level, if the user were to publish the extended public key and others were to derive the same faulty child key, the security of the parent key would be degraded.
 
-For example, an implementer could mark the index as forbidden and simply increment the child index.
-However, depending on the level, should the user share their extended public key to the world and others derive the same faulty child key, the security of the parent key would be degraded.
-
-Regardless, the likelihood of either happening is highly unlikely.
-
-\subsection{The 28 left bytes of Z}
-\label{subsec:the_28_left_bytes_of_z}
+Fortunately, it is extremely unlikely that either case will happen, based on the security of the hash function used for derivation.
 

### 2024-05-23 paper.tex
clean up technical description of BIP32-Ed25519

@@
 
-As was mentioned in \cref{subsec:elliptic_curves_and_their_key_spaces}, there is a need to ensure that bits 0, 1, 2, 3 and 255 are 0 and that bit 254 is 1 for private keys (scalars) in Ed25519.
+\subsection{The 28 bytes of Z}
+\label{subsec:the_28_bytes_of_z}
 
-The paper's method for ensuring this is to produce the child private key respectively public key as follows:
+As was mentioned in \cref{subsec:elliptic_curves_and_their_key_spaces}, for Ed25519 private-keys scalars, one must ensure that bits 0, 1, 2, 3 and 255 are 0, and that bit 254 is 1.
+The paper's method for ensuring this is to produce the child private key and public key as follows.
 
-Derive $Z$ from either the parent's public key or parent's extended private key, depending on whether the child is hardened or not:
+Derive $Z$ from either the parent's public key $A^{P}$ or parent's private key $k^{P}$, depending on whether the child is hardened or not:
+\[ Z \leftarrow
+  \begin{cases} 
+    F_{c^P}(\mathtt{0x02} \| A^P \| i) & i \in  [0, 2^{31}) \\
+    F_{c^P}(\mathtt{0x00} \| k^P \| i) & i \in [2^{31}, 2^{32}) .
+  \end{cases} \]
+Then truncate the 32-byte $Z$ to 28 bytes (224 bits): $Z_L \leftarrow Z_{[:224]}$.
 
-$  Z \leftarrow 
-  \begin{cases} 
-  F_{c^P}(0x02||A^P||i) & i \in  [0, 2^{31}) \\
-  F_{c^P}(0x00||k^P||i) & i \in [2^{31}, 2^{32})
-  \end{cases}
-  $
+Depending on hardening, calculate the child private key (scalar) or public key (which corresponds to an address on the blockchain):
+\begin{align*}
+  k_L &\leftarrow \langle 8[Z_L]  + [k_L^P]\rangle \\
+  A_i &\leftarrow A^P + [8Z_L]B .
+\end{align*}
+Here~$B$ is the base point of the curve, and~$i$ is a little-endian 4-byte string.
 
-Truncate the 32-byte $Z$ to 28 bytes (224 bits): $Z_L \leftarrow Z_{[:224]}$.
+The calculations of~$k_R$ (which is necessary to produce the full private key $k = (k_L, k_R)$) and the child chain code $c_i$ (which is necessary to serve as the key for the HMAC-512 function~$F$, and is part of the extended public key $(A_i, c_i)$) have been omitted for brevity.
 
-Depending on the choice, calculate the child private key (scalar) or public key (which corresponds to an address on the blockchain):
-
-\begin{enumerate}[label={}] 
-  \item $k_L \leftarrow \langle 8[Z_L]  + [k{_L}^P]\rangle$ 
-  \item $A_i \leftarrow A^P + [8Z_L]B$
-\end{enumerate}
-
-The calculations of $k_R$, which are necessary to produce the extended private key $k = (k_L, k_R)$, and the child chain code $c_i$, which is necessary to serve as the key for the HMAC-512 function $F$ (and be part of the extended public key $(A_i, c_i)$); have been omitted for brevity.
-$B$ represents the base point of the curve and $i$ a little-endian 4-byte string.
-
-By truncating $Z$ to 28 bytes and then multiplying by 8, the authors ensure that the scalar is a multiple of 8 and that bit 255 is 0. When $k{_L}^P$ is added, it contributes a 1 on bit 254.
+By truncating $Z$ to 28 bytes and then multiplying by 8, the authors ensure that the scalar is a multiple of 8 and that bit 255 is 0.
+When $k_L^P$ is added, it contributes a 1 at bit 254.
 
@@
 This prevents overflow when adding, which would affect bit 254.
+As a result, the child private-key scalars are guaranteed to meet the requirements of Ed25519, for many levels of derivation.
 
-As a result, the children are guaranteed to be safe in accordance with the requirements of Ed25519, for many levels of derivation.
-
-To understand how many levels can be safely performed, consider that the root key $a$ can be expressed as follows: $a = 2^{254} + 8b$, where $b < 2^{250}$, due to the bit manipulation and the way it is generated.
-
-Thus, $a = 2^{254} + 8b < 2^{254} + 2^{253}$.
+To understand how many levels can be safely performed, consider that the root key~$a$ can be expressed as $a = 2^{254} + 8b < 2^{254} + 2^{253}$, where $b < 2^{250}$ due to the bit manipulation and the way it is generated.
 
@@
 
-In short, the key at level $D$ depth (called $j$ in the paper) is defined by the inequality, $a_D < a + D\cdot2^{227} = 2^{254} + 2^{253} + D\cdot2^{227}$.
+In short, the key at depth~$D$ (called~$j$ in the paper) satisfies the inequality $a_D < a + D\cdot2^{227} < 2^{254} + 2^{253} + D\cdot2^{227}$.
 
-The paper specifies $D \leq 2^{20}$, allowing for a maximum 1048576 levels of derivation, such that $a_{D} \leq 2^{254} + 2^{253} + 2^{227 + 20 = 247}$ - a form suitable for our purposes.
+The paper specifies $D \leq 2^{20}$, allowing for a maximum 1048576 levels of derivation, so that $a_{D} < 2^{254} + 2^{253} + 2^{247}$, a form that is compatible with the Ed25519 requirements.
 
-Unfortunately, offering this many levels comes at a cost: only 28 bytes (224 bits) of randomization takes place for every level.
+Unfortunately, offering this many levels comes at a cost: only 28 bytes (224 bits) of randomization takes place for each level.
 
 Consider the scenario where a single child private key is compromised.
-The security of its parent key would degrade to around only $\sqrt{2^{224}}=2^{112}$ due to the Birthday Attack, below the 128 bits of security offered by Ed25519.
-This contradicts the author's claims that no loss of security occurs as a result of their method.
+The security of its parent key would degrade to around only $\sqrt{2^{224}}=2^{112}$ due to standard rho/Kangaroo attacks, below the 128 bits of security offered by Ed25519.
+This contradicts the authors' claims that no loss of security occurs as a result of their method.
 

### 2024-05-23 paper.tex
clean up and correct description of our improvements

@@
 \usepackage{enumitem}
+\usepackage{booktabs}
 
@@
 
-The purpose of our modifications is to make derived keys less related, so that there is less security degradation if some private keys are compromised.
+The purpose of our modifications is to make derived keys more randomized and less related, so that there is less security degradation if some private keys are compromised.
 
-Define $g$, such that $Z_L = Z_{[:256-g]}$. The BIP32-Ed25519 paper~\cite{BIP32-Ed25519} set $g=32$ to arrive on 28 bytes.
+We introduce a positive integer parameter~$g$, and instead define $Z_L = Z_{[:256-g]}$ to be all but~$g$ bits of~$Z$.
+The BIP32-Ed25519 paper~\cite{BIP32-Ed25519} uses $g=32$, so that $Z_{L}$ is 28 bytes.
 
-An Ed25519 scalar has 251 bits available for randomization (255, 254, 2, 1, 0 are fixed), with an additional bit (253) having to be 0 to avoid overflow.
-The level at which we can no longer guarantee 253 will be 0 forms an upper bound.
+An Ed25519 scalar has 251 bits available for randomization (bits 255, 254, 2, 1, 0 are fixed), i.e., it must be a multiple of 8 in the range $[2^{254},2^{255})$.
+Additionally,~\cite{BIP32-Ed25519} requires that the master scalar have bit 253 set to~0 to avoid overflow in the children, i.e., it must be less than $2^{254} + 2^{253}$.
+The level at which we can no longer guarantee that a child scalar meets the Ed25519 requirements determines the depth~$D$ (i.e., number of derivation levels) we can handle.
 
-Recall from \cref{subsec:the_28_left_bytes_of_z} that each $Z_L < 2^{256 - g}$ and is multiplied by 8.
-Root $a$ must also be less than $2^{254} + 2^{253}$.
-Define $d = log_2(D)$.
-The previous paragraph can thus be expressed as follows:
+Recall from \cref{subsec:the_28_bytes_of_z} that each $Z_L < 2^{256 - g}$ and is multiplied by 8, so all derived private-key scalars are multiples of 8.
+The root scalar $a < 2^{254} + 2^{253}$.
+Define $d = \log_2(D)$, where~$D$ is the number of derivation levels.
+Then a bound on any derived private-key scalar~$a'$ is
 \begin{align*}
-  a_D &< a + 8\cdot D\cdot2^{256 - g} \\
-  a_D &< a + D\cdot2^{259 - g} \\
-  a_D &< a + 2^{259 + d - g}\\
-  a_D < 2^{254} +& 2^{253} + 2^{259 + d - g} < 2^{254} + 2^{254}
+  a' &< a + 8\cdot D\cdot2^{256 - g} \\
+     &= a + D \cdot 2^{259 - g} \\
+     &= a + 2^{259 + d - g} \\
+     &< 2^{254} + 2^{253} + 2^{259 + d - g} .
 \end{align*}
-The relationship between $d$ and $g$ can be stated as:
-\begin{align*}
-  253 &> 259 + d - g \\
-  &\Downarrow \\
-  g &> d + 6
-\end{align*}
-
- $g = d + 6$ defines the first level at which the 3rd highest bit overflows into the 2nd highest bit, which is set to 1 already and would overflow into the highest bit.
-$d$ must necessarily be $d < g - 6$.
-
- As a result, the number of safe levels $S$ of derivation that can be performed is one level fewer than $D = 2^{d} < 2^{g - 6}$:
-
- \begin{equation*}
-  S = D - 1 = 2^{d} - 1
-  \end{equation*}
+So, in order to ensure that $a' < 2^{255}$, it suffices to set $g,d$ so that
+\[ g \geq d + 6 .
+\]
+By the above, this implies that $a' < 2^{254} + 2^{253} + 2^{253} = 2^{255}$, as needed.
 
@@
   \centering
-  \begin{tabular}{|c|c|c|c|}
-  \hline
-  g & d  & D & S  \\
-  \hline
-  8 & 2 & 4 & 3 \\
-  9 & 3 & 8 & 7 \\
-  16 & 10 & 1024 & 1023 \\
-  32 & 26 & 67108864 & 67108863  \\
-  \hline
+  \begin{tabular}{ccc}
+    \toprule
+    $g$ & $d$ & $D$ \\
+    \midrule
+    8 & 2 & 4 \\
+    9 & 3 & 8 \\
+    16 & 10 & 1024 \\
+    32 & 26 & 67108864 \\
+    \bottomrule
   \end{tabular}
-  \caption{Bits to randomize $g$ resulting in levels $D$. $S$ is the number of safe levels.}
-  \label{tab:g_and_d}
+  \caption{Number of bits~$g$ to randomize, and corresponding number of safe derivation levels~$D$.}\label{tab:g_and_d}
 \end{table}
@@
 \Cref{tab:g_and_d} presents a selection of possible values.
-$g=8$ is the most secure and breaks at the byte boundary (256 bits - 8 bits = 31 bytes) but it results in only 3 levels of derivation - too few to satisfy the derivation schemes mentioned in \cref{subsec:hd_wallets}, which require 5.
+While $g=8$ is the most secure among these options, and breaks~$Z$ at the byte boundary ($256-8$ bits is 31 bytes), it gives only $D=4$ levels of derivation, which is too few to satisfy the derivation schemes mentioned in \cref{subsec:hd_wallets} (which require 5).
 
-$g=16$ is the next value that breaks at the byte boundary (equal to 30 bytes) and provides more than enough levels. It randomizes 240 out of the 251 bits available in an Ed25519 scalar.
+Similarly, $g=16$ is the next value that breaks at the byte boundary (30 bytes) and provides more than enough levels.
+However, it randomizes only 240 out of the 251 bits available in an Ed25519 scalar.
 
-$g=9$ is the most secure value that also allows for enough derivation levels, allowing for 7 levels of derivation.
-It randomizes 246 out of the 251 bits available in a Ed25519 scalar.
-The only downside is that it does not break at the byte boundary, requiring a bit of extra care when implementing.
+Finally, $g=9$ is the most secure value that also allows for enough derivation levels: up to 8.
+It randomizes 247 out of the 251 bits available in a Ed25519 scalar.
+A downside is that it does not break~$Z$ at the byte boundary, requiring a bit of extra care when implementing.
 
-It also becomes apparent that the recommended maximum level of depth in the BIP32-Ed25519 paper, $2^{20}$, is lower than the actual number of levels that can be performed without bit overflows breaking the scalar's Ed25519-compatibility: $2^{26} - 1$.
-
-Indeed, for $D=2^{20}$, $a_{D} \leq 2^{254} + 2^{253} + 2^{247} < 2^{254} + 2^{253} + 2^{227+26=253} = 2^{255}$.
+It also becomes apparent that the quoted maximum depth~$2^{20}$ from~\cite{BIP32-Ed25519} is smaller than the actual number of safe levels, which is at least $2^{26}$.
 

### 2024-06-13 paper.tex
citation sort, title, author
remove cite from author line
touch up abstract

@@
-\title{Bip32-Ed25519++: A More Secure Hierarchical Deterministic Wallet Standard for Ed25519 Keys}
+\title{BIP32-Ed25519++: A More Secure Hierarchical Deterministic Wallet Standard for Ed25519 Keys}
 
-% \author{Chris Peikert\cite{ChrisP}}
+\author{Chris Peikert}
 
@@
     \begin{abstract}
-      \abstractText
-      Implementing BIP32-style HD wallets for Ed25519 keys is a non-trivial task due to the non-linear key space of Ed25519.
-The paper BIP32-Ed25519 Hierarchical Deterministic Keys over a Non-linear Keyspace has seen adoption in the industry, but has some security issues.
-In this paper modifications are proposed that addresses those issues, at the cost of reducing the number of possible levels of child keys.
-      \newline
-      \newline
+      \abstractText Implementing BIP32-style Hierarchical Deterministic (HD) wallets for Ed25519 keys is a non-trivial task, due to the non-linear key space of Ed25519.
+      A proposal by Khovratovich and Law has seen adoption in the industry, but it has some security issues.
+      This paper proposes some small modifications that address those issues, at the cost of reducing the number of levels in the hierarchy.
+      \newline \newline
     \end{abstract}
@@
 \nocite{*}
-\bibliographystyle{plain}
+\bibliographystyle{unsrt}
 \bibliography{sources}

### 2024-06-13 paper.tex
clarify "level"/hardening for indices

@@
 
-Each level has $2^{32}$ possible values.
-When hardening a level, the most-significant bit is set to 1.
-Thus $[0, 2^{31})$ are reserved for non-hardened keys, and $[2^{31}, 2^{32})$ are reserved for hardened keys.
+There are $2^{32}$ possible values for the address index (\verb=key=).
+In a hardened index, the most-significant bit is set to 1.
+So, the interval $[0, 2^{31})$ is for non-hardened keys, and $[2^{31}, 2^{32})$ is for for hardened keys.
 

### 2024-06-13 paper.tex
minor wording/clarity changes

@@
 
-In the basic case, a single seed phrase corresponds to a single address, requiring users to keep track of many seed phrases should they wish to use separate addresses for different application domains: making and receiving payments, holding NFTs, participating in a DAO, DeFi, gaming, etc.
+In the basic case, a single seed phrase corresponds to a single address, requiring users to keep track of many seed phrases should they wish to use separate addresses for different application domains: e.g., making and receiving payments, holding NFTs, participating in a DAO, DeFi, gaming, etc.
 
@@
 A new public key can be formed by adding $[x]B$ to the public key $[k]B$, where $x$ and $k$ are private keys (scalars), and $B$ is the base point.
-This results in a new public key whose private key is $[x] + [k] \bmod q$, where $q$ is the order of the field.
+This results in a new public key whose private key is $[x] + [k] \bmod q$, where $q$ is the order of the scalar field.
 
@@
 
-Unfortunately, private keys in Ed25519 have to be derived by hashing a private ``seed'' and manipulating the bits.
+However, private keys in Ed25519 have to be derived by hashing a private ``seed'' and manipulating the bits.
 RFC\-8032~\cite{RFC8032} specifies that in the 32-byte private key:
@@
 \end{itemize}
-So, by definition, the key space for Ed25519 is non-linear relative to the group order.
+So, by definition, the key space for Ed25519 is non-linear relative to the group order: the sum of two private keys (modulo the group order) is not necessarily a valid private key.
 
@@
 
-In this paper an improvement to their method is proposed that provides a more secure way to derive Ed25519 keys, at the cost of reducing the number of levels of child keys.
+This paper proposes an improvement to the method from~\cite{BIP32-Ed25519} that provides a more secure way to derive Ed25519 keys, at the cost of reducing the number of levels in the hierarchy of child keys.
 
@@
 
-While reviewing the BIP32-Ed25519 paper~\cite{BIP32-Ed25519}, two issues were identified, the first minor and the other major.
+While reviewing the BIP32-Ed25519 paper~\cite{BIP32-Ed25519}, two issues were identified, the first minor and the other significant.
 
@@
 
-While deriving a new private or public child key, it is possible to produce a child that is unsafe to use, in the following two cases:
+When deriving a new private or public child key, it is possible to produce a child that is unsafe to use, in the following two cases:
 \begin{itemize}[label=\textendash, itemsep=-0.5em]
@@
 
-As was mentioned in \cref{subsec:elliptic_curves_and_their_key_spaces}, for Ed25519 private-keys scalars, one must ensure that bits 0, 1, 2, 3 and 255 are 0, and that bit 254 is 1.
-The paper's method for ensuring this is to produce the child private key and public key as follows.
+As mentioned in \cref{subsec:elliptic_curves_and_their_key_spaces}, for Ed25519 private-keys scalars, one must ensure that bits 0, 1, 2, 3 and 255 are 0, and that bit 254 is 1.
+The method proposed in~\cite{BIP32-Ed25519} for ensuring this is to produce the child private key and public key as follows.
 
@@
 
-To understand how many levels can be safely performed, consider that the root key~$a$ can be expressed as $a = 2^{254} + 8b < 2^{254} + 2^{253}$, where $b < 2^{250}$ due to the bit manipulation and the way it is generated.
+To understand how many levels can be safely derived, consider that the root key~$a$ can be expressed as $a = 2^{254} + 8b < 2^{254} + 2^{253}$, where $b < 2^{250}$ due to the bit manipulation and the way it is generated.
 
@@
 
-The purpose of our modifications is to make derived keys more randomized and less related, so that there is less security degradation if some private keys are compromised.
+The purpose of our modifications is to add more randomization to child keys, so that there is less security degradation if some private keys are compromised.
 
@@
 
-Similarly, $g=16$ is the next value that breaks at the byte boundary (30 bytes) and provides more than enough levels.
+Similarly, $g=16$ is the next value that breaks at the byte boundary (30 bytes), and it provides more than enough levels.
 However, it randomizes only 240 out of the 251 bits available in an Ed25519 scalar.
@@
 
-It also becomes apparent that the quoted maximum depth~$2^{20}$ from~\cite{BIP32-Ed25519} is smaller than the actual number of safe levels, which is at least $2^{26}$.
+From this more general setup it is also apparent that the quoted maximum depth~$2^{20}$ from~\cite{BIP32-Ed25519} is smaller than the actual number of safe levels, which is at least $2^{26}$.
 
@@
 
-In particular, setting $g=9$ allows for the most secure implementation that also satisfies the requirements of the most common derivation schemes.
+In particular, setting $g=9$ allows for the most secure implementation that also satisfies the requirements of Ed25519 and the most common derivation schemes.
 
